---
title: "CodeBook"
author: "Timo Ruohomäki"
date: "2023-07-16"
output:
  md_document:
    variant: markdown_github
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting and Cleaning Data Course Project

The purpose of this project is to demonstrate your ability to collect, work with, and clean a data set. The goal is to prepare tidy data that can be used for later analysis.

The assignment deliverable is a R script that does the following: 

1) Merges the training and the test sets to create one data set.

2) Extracts only the measurements on the mean and standard deviation for each measurement. 

3) Uses descriptive activity names to name the activities in the data set

4) Appropriately labels the data set with descriptive variable names. 

5) From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

This codebook has been created using with *Knitr* using *rmarkdown* package.

## Description of Variables

The following code is used to import the various datasets included in the dataset as separate text files.
The variables are:

| variable | description |
|----|----|
| featureData | List of all features |
| activityLabels | Links the class tables with their activity name |
| subjectTest | Each row identidies the subject who performed the activity for each window sample, with a range from 1 to 30 |
| xTestData | Test set |
| yTestData | Test labels |
| subjectTrain | Each row identidies the subject who performed the activity for each window sample, with a range from 1 to 30 |
| xTrainData | Train set |
| yTrainData | Train labels |

In the source dataset features are normalized and bounded within [-1,1].
Each feature vector is a row on the text file

For more details on variables within the data sets, see features_info.txt that has been included in the dataset.

## Dependencies (Packages)

The following packages are required to run the script:

```{r}
library(readr)
library(knitr)
library(dplyr)
```


## Step 1: Merge the training and the test sets to create one data set.

```{r Step 1, echo=TRUE}
  
  featureData <- read.table("ucidataset/features.txt", col.names = c("n","functions"))
  activityLabels <- read.table("ucidataset/activity_labels.txt", col.names = c("code","activity"))
  subjectTest <- read.table("ucidataset/test/subject_test.txt", col.names = "subject")

  xTestData <- read.table("ucidataset/test/X_test.txt", col.names = featureData$functions)
  yTestData <- read.table("ucidataset/test/Y_test.txt", col.names = "code")

  subjectTrain <- read.table("ucidataset/train/subject_train.txt", col.names = "subject")

  xTrainData <- read.table("ucidataset/train/X_train.txt", col.names = featureData$functions)
  yTrainData <- read.table("ucidataset/train/Y_train.txt", col.names = "code")

  xData <- rbind(xTrainData, xTestData)
  yData <- rbind(yTrainData, yTestData)

  subjectData <- rbind(subjectTrain, subjectTest)
  
  # create one master (merged) dataset

  masterDataset <- cbind(subjectData, xData, yData)
  
  str(masterDataset)

```


## Step 2: Extract only the measurements on the mean and standard deviation for each measurement

```{r echo=TRUE}
  
  extractDataset <- masterDataset %>% select(subject, code, contains("mean"), contains("std"))

  
```

## Step 3: Use descriptive activity names to name the activities in the data set

```{r Step 3}

  extractDataset$code <- activityLabels[extractDataset$code, 2]

```

## Step 4: Appropriately label the data set with descriptive variable names

```{r}

names(extractDataset)[1] = "Subject"
names(extractDataset)[2] = "Activity"
names(extractDataset) <- gsub("Acc", "Accelerometer", names(extractDataset))
names(extractDataset) <- gsub("BodyBody", "Body", names(extractDataset))
names(extractDataset) <- gsub("Gyro", "Gyroscope", names(extractDataset))
names(extractDataset) <- gsub("Mag", "Magnitude", names(extractDataset))
names(extractDataset) <- gsub("^t", "Time", names(extractDataset))
names(extractDataset) <- gsub("^f", "Frequency", names(extractDataset))
names(extractDataset) <- gsub("tBody", "TimeBody", names(extractDataset))
names(extractDataset) <- gsub("-mean()", "Mean", names(extractDataset), ignore.case = TRUE)
names(extractDataset) <- gsub("-std()", "STD", names(extractDataset), ignore.case = TRUE)
names(extractDataset) <- gsub("-freq()", "Frequency", names(extractDataset), ignore.case = TRUE)
names(extractDataset) <- gsub("angle", "Angle", names(extractDataset))
names(extractDataset) <- gsub("gravity", "Gravity", names(extractDataset))

```


## Step 5: From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

```{r}

dataProduct <- extractDataset %>%
  group_by(Subject, Activity) %>%
  summarise_all(funs(mean))

write_excel_csv(dataProduct, "UCI-DataProduct.csv")

```


